Cultural Transmission and the Social Contexts of Technological Transitions during the Late Paleolithic of Korea

### 2023-04-13 analysis/paper/paper.qmd
Edited in response to Fitzhugh's Fitzhugh's comments

One limitation of this aggregation of stemmed points into chronological phases is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points. These isolated finds are ambiguous with respect to local traditions of artifact making, so to further investigate temporal change, we focus only on the four assemblages that have more than five stemmed points (@fig-plot-cor-four-assemblage). We assume these four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making. The four assemblages include three from Suyanggae (SYG) and one from Yongsandong (YS). @fig-plot-cor-four-assemblage shows that at this finer scale of resolution the picture is complex. Phase 2 shows both highly correlated attributes, from SYG1_2, and relatively uncorrelated attributes, from SYG6_2. The two assemblages from Phase 3 show similarly highly correlated attributes, with the median right around our threshold value of `r threshold_value_cor`; however, the spread of correlations is high, suggesting a mixture of transmission biases.

@@

The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological Phase 3, after 25 ka. It is in the same range as the correlation coefficients that Bettinger and Eerkens interpreted as indirect bias. We also found a slight decrease in CV values for the four attributes. Those results suggest a shift away from trial and error and towards greater reliance on copying a model in the social context of tool-making skills transmission. Drawing on evolutionary ecological theory, we might speculate that the cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf. @fitzhugh2001risk]. With more frequent contacts and connections with members of social networks, there may have been more opportunities to learn point-making directly from socially successful individuals.

@@

We concede that our work is rather speculative and provides only a limited explanation of the social contexts surrounding the spread of stemmed points. Although social learning and migration or drift models requires approximate estimates for population and generation turnover required to transmit knowledge of new technology, we were unable to employ this approach due to the large number of assumptions required to cover a 25,000-year time span. In her simulation study, Garvey was able to derive the population at a particular site based on the number of rooms [@garvey2018current], but our research sites lacked sufficient information to utilize this method. Moreover, we recognize that we did not account for the possibility of the shape of stone artifacts being modified through reuse and resharpening, which can result in the remaining shape of stemmed points being different from their original form through inherited technology (i.e. the Frison Effect) [@dibble1995middle; @jelinek1976form; @frison1968functional]. Furthermore, we only considered two types of biases, as our primary goal was to understand the general trajectory of technological transition, and we had limited empirical evidence to explore a multiplicity of conjectural scenarios. In future research, we plan to focus more closely on individual assemblages and site occupation to gain a better understanding of the population responsible for creating and using stemmed points. By so doing, we anticipate achieving a more exhaustive understanding of the morphology of stemmed points and the mechanisms behind their transmission.

>>>>

One limitation of this aggregation of stemmed points into chronological phases is that the sample consists of a relatively large number of assemblages with only 1-2 stemmed points. These isolated finds are ambiguous with respect to local traditions of artifact making, so to further investigate temporal change, we focus only on the four assemblages that have more than five stemmed points (@fig-plot-cor-four-assemblage). We assume these four assemblages are more likely to represent a consistent, recurring way of making stemmed points than isolated finds, and thus more relevant for comparing modes of cultural transmission of artifact making. The four assemblages include three from Suyanggae (SYG) and one from Yongsandong (YS). @fig-plot-cor-four-assemblage shows that at this finer scale of resolution the picture is complex. Phase 2 shows both highly correlated attributes, from SYG1_2, and relatively uncorrelated attributes, from SYG6_2. The two assemblages from Phase 3 show similarly highly correlated attributes, with the median right around our threshold value of `r threshold_value_cor`; however, the spread of correlations is high, suggesting some influence of indirect bias.

@@

The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological Phase 3, after 25 ka. It is in the same range as the correlation coefficients that Bettinger and Eerkens interpreted as indirect bias. We also found a slight decrease in CV values for the four attributes. Those results suggest a shift away from trial and error and towards greater reliance on copying a model in the social context of tool-making skills transmission. Drawing on evolutionary ecological theory, we might speculate that the cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf. @fitzhugh2001risk]. Another possibility is that the cooler climate might result in better overall fitness due to an increase in available food resources, or an increase in the patchiness of food and water resources, which would have concentrated them in fewer predictable locations. In either case, these conditions might lead to larger networks of individuals interacting with one another and an increased ability to learn from experienced social teachers or successful models.

@@

We concede that our work is rather speculative and provides only a limited explanation of the social contexts surrounding the spread of stemmed points. Although social learning and migration or drift models requires approximate estimates for population and generation turnover required to transmit knowledge of new technology, we were unable to employ this approach due to the large number of assumptions required to cover a 25,000-year time span. In her simulation study, Garvey was able to derive the population at a particular site based on the number room block (house) and assumed four learning generation for about 100 years of site occupation [@garvey2018current], but our research sites lacked sufficient information to utilize this method. In our future study, we will further investigate how to quantify multiple variables and conduct simulations to develop a logical bridge between our primary approach borrowed from Bettinger and Eerkens [-@bettinger1999point] and social contexts for the appearance of stemmed points. We acknowledge that we did not account for the possibility of the shape of stone artifacts being modified through reuse and resharpening, which can result in the remaining shape of stemmed points being different from their original form through inherited technology (i.e. the Frison Effect) [@dibble1995middle; @jelinek1976form; @frison1968functional]. It is our assumption that the majority of stemmed points used in our research have retained their original shape, given that retouch is rare on the tip of stemmed points. In our study, we only considered two types of biases as social contexts without considering other factors that impact on morphology of stemmed points such as tool efficiency and manufacturing economization. Mainly due to limited empirical evidence to explore a multiplicity of conjectural scenarios, we decided to focus on our primary goal, understanding the general trajectory of technological transition. In future research, we plan to focus more closely on individual assemblages and site occupation to gain a better understanding of the population responsible for creating and using stemmed points. By so doing, we anticipate achieving a more exhaustive understanding of the morphology of stemmed points and the mechanisms behind their transmission.
